Practical consistency management for geographically distributed MMOG servers

### 2011-05-27 textos/optpaxos/draftv2.tex
propPending

To ensure reliability despite server crashes, each server is replicated using state-machine replication, implemented with Paxos. Each player sends his commands to the server to which he is connected, which then forwards it to the coordinator. Each command is assigned a timestamp and executed against objects in timestamp order. We implement this by using a logical clock in each server group. Guaranteeing that the same set of commands is executed by the respective affected objects in the same order provides the level of consistency we are seeking. Therefore, the challenge is how to assign timestamps to commands such that consistency is not violated and commands are not discarded due to stale timestamp values.

@@

Instead of having the conservative delivery done completely independently from the optimistic one, we can actually use the latter as a hint for the final delivery order. Since the optimistic delivery should be fast when compared to the conservative one, waiting for it should not decrease the system performance significantly. Also, if we wait a short period longer, we can avoid a rollback later caused by mismatches between the two delivery orders, which is not desirable. The basic idea is, then, to pick the optimistic delivery order seen at the coordinator of each group.

@@

There are three possibilities for each message $m$, in the perspective of the coordinator $c$ of a given group $G$:

@@

  In this case, when $m$ is optimistically delivered by the coordinator $c$, $c$ also atomically broadcasts $m$ in $G$. When it is atomically delivered by a process $p$ of group $G$, it is not immediately delivered to the application. Instead, it is inserted in a \textit{consPending} list for later being conservatively delivered, once every group $G'$ in $sendersTo(G)$ have already sent a message $barrier(G',G) = t$, such that $t > m.ts$. This is done because there could be a message $m'$ yet to come from another group $G'$, such that $m'.ts < m.ts$.

@@

  In this case, when $m$ is optimistically delivered by the coordinator, it is also atomically broadcast, which is done by running a Paxos consensus instance within $G$, but having as learners the processes of the groups in $m.dst$. The proposed value is $\{m, barrier(G, G')\}$, where $barrier(G,G')$ is simply the value of $m.ts$. When the consensus is achieved, both the message and the barrier from $G$ are learnt by the processes of every destination group $G'$. The processes of all the involved groups also insert $m$ into the respective \textit{consPending} list.

@@

  In this case, when $m$ is optimistically delivered by a process $p$ of group $G$, nothing else is done. The message $m$ should be inserted into the \mbox{\textit{consPending}} list of $p$ only when learnt via a consensus instance started at its group of origin -- stored in $m.src$.
\end{itemize}

>>>>

To ensure reliability despite server crashes, each server is replicated using state-machine replication, implemented with Paxos. Each player sends his commands to the server to which he is connected, which then proposes that command in a consensus instance. Each command is assigned a timestamp and executed against objects in timestamp order. We implement this by using a logical clock in each server group. Guaranteeing that the same set of commands is executed by the respective affected objects in the same order provides the level of consistency we are seeking. Therefore, the challenge is how to assign timestamps to commands such that consistency is not violated and commands are not discarded due to stale timestamp values.

@@

Instead of having the conservative delivery done completely independently from the optimistic one, we can actually use the latter as a hint for the final delivery order. Since the optimistic delivery should be fast when compared to the conservative one, waiting for it should not decrease the system performance significantly. Also, if we wait a short period longer, we can avoid a rollback later caused by mismatches between the two delivery orders, which is not desirable. The basic idea is, then, to pick the optimistic delivery order seen at the processes of each group.

@@

We use consensus to conservatively deliver each message. Consider that each consensus instance $I$ from each group $I.grp$ receives a monotonically increasing unique integer identifier, without gaps, that is, for any two instances $I_i$ and $I_k$, such that $I_i.grp = I_k.grp$, if $I_i.id + 1 < I_k.id$, there is necessarily an instance $I_j : I_i.grp = I_j.grp = I_k.grp \wedge I_i.id < I_j.id < I_k.id$. Consider also that each message sent by a group $G$ has a group sequence number $m.gs$ related to the order in which it is conservatively delivered, relatively to other messages also sent by $G$. As messages are conservatively delivered via consensus, the group sequence number of a message is equal to the id of the instance in which it was decided, that is, $m.gs = i \Leftrightarrow \exists I : I.grp = m.src \wedge Decide(I.id, m) \wedge I.id = i$.

There are three possibilities for each message $m$, in the perspective of a process $p$ of $G$:

@@

  In this case, when $m$ is optimistically delivered by $p$, $p$ checks whether the latest consensus instance $I_{prev}$ in which it participated, or is trying to start, has already been decided -- if not, $p$ enqueues $m$ in a queue $propPending$ as the next message being proposed by it, so other tasks can keep being executed. Then, once $I_{prev}$ has been decided, $p$ checks wheter $m$ -- or a message $m' : m'.ts > m.ts$ -- has already been conservatively delivered in some previous consensus instance. If such message was indeed delivered previously, $p$ discards $m$; otherwise $p$ proposes $m$ as the next message to be conservatively delivered by the processes of $G$. When $m$ is conservatively delivered, it is not immediately delivered to the application. Instead, it is inserted in a $consPending$ list for later being conservatively delivered, which will happen once every group $G'$ in $sendersTo(G)$ have already sent a message $barrier(G',G) = t$, such that $t > m.ts$. This is done because there could be a message $m'$ yet to come from another group $G'$, such that $m'.ts < m.ts$.

@@

  In this case, when $now > m.ts + w(p)$, $p$ tries to initiate a consensus instance within $G$ to decide $m$. If $p$ cannot start the proposal now, $m$ is enqueued in $propPending$ for being proposed later. Before proposing $m$, $p$ checks whether $m$ -- or a message $m' : m'.ts > m.ts$ -- has already been decided in some previous consensus instance. If it has, $m$ is discarded; otherwise, $p$ starts a consensus instance proposing $m$. Once any message $m$ is decided, if $G \in m.dst$, $m$ is inserted in the $consPending$ list. Besides, $p$ sends a message $\{m, \text{`cons'}\}$ to every $p' : (\exists G' \in m.dst \setminus \{G\}) \wedge (p' \in G')$. When $\{m, \text{`cons'}\}$ is received by each $p' \in G'$, $p'$ checks whether it has ever inserted $m$ in its own $consPending$ list. If not, $p'$ inserts $m$ into $consPending$ and adjusts $barrier(G, G')$ to $m.ts$.

@@

  In this case, when $m$ is optimistically delivered by a process $p$ of group $G$, nothing else is done. The message $m$ should be inserted into the \mbox{\textit{consPending}} list of $p$ only when some message $\{m, \text{`cons'}\}$ is received.
\end{itemize}
